---
title: "Experimental Design for Automatic Egg Incubators Using a 2^2 factorial Design
  and RCBD."
author: "STAT.BONVENTURE.M.MIRUKA"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## OBJECTIVE:

To determine whether Temperature, Humidity or their interaction is the primary 
factor driving variations in egg hatchability rate (Y) and quantify
individual-level incubator variability.

## Experimental design:

We utilize a 2^2 full Factorial Design in a Randomized Complete Block Design.

Factor A : Temperature(T)
           Low(To): standard low (36.5℃)
           High(T1): standard optimal(37.8℃)
           
Factor B : Relative Humidity(H)
           Low(Ho): standard low(45% RH)
           High(H1): standard optimal(55% RH)
           
Blocks : Incubator models - M1, M2, M3, M4

Experimental units: Fertilized eggs.

Replications : r=3runs per model treatment combination.



## Statistical analysis framework

Model : Two-way ANOVA with blocking
Model equation : Yijk= μ + αi + βj + (αβ)ij + γk + εijk, where
                            
                            μ : overall mean hatchability rate
                            αi: main effect of temp level i
                            βj: main effect of humidity level j
                            (αβ)ij: interaction effect
                            γk: blocking effect of model k
                            εijk: random error ~N(0,σ^2)
                            

## Dataset generation

```{r}
set.seed(42)
incubator_data= data.frame(incubator_model= factor(rep(c("M1","M2","M3","M4"),
                                                       each =12)),
                           temp_level=
                             factor(rep(rep(c("low","high"),each=6),4),levels=c("low","high")),
                           humidity_level= factor(rep(rep(c("low","high"), each=3),8),levels=c("low","high")),
                           hatchability_rate= c(62.4,60.8,63.1,68.5,67.2,69.0,74.2,73.0,75.1,88.4,89.1,87.6,58.2,59.0,57.5,64.1,63.5, 65.0,70.5,71.2,69.8,83.0,84.2,82.5,51.0,49.5,53.2,56.8,58.0,55.5,64.0,62.5,65.1,72.4,74.0,71.8,64.5,63.0,65.2,71.0,70.2,72.1,77.8,76.5,78.2,91.2,92.5,90.8)
                           )
head(incubator_data)
```

## ANOVA 


### libraries

```{r}
library(stats)
```

### Step 1 :Two_way anova with blocking (RCBD).

```{r}
anova_model=aov(hatchability_rate~ temp_level * humidity_level + incubator_model,data=incubator_data)

cat("=== TWO_WAY ANOVA TABLE===\n")
summary_anova= summary(anova_model)
print(summary_anova)
```
NOTE: by the view of the Anova table p values of extremely less than 0.05 indicate that there is significant effect of the various factors towards hatchability rates.


### Step 2: Calculating effect size (eta-squared) to identify primary factor.


```{r}
ss= summary_anova[[1]]["Sum Sq"]
eta_squared= ss/sum(ss)
rownames(eta_squared)=trimws(rownames(ss))
colnames(eta_squared)="Eta_Squared"

cat("\n===EFFECT SIZE (ETA-SQUARED/ PROPORTION OF VARIANCE)===\n")
print(round(eta_squared,4))
```
NOTE: -Temperature levels account for the vast majority of total variance 
       confirming it as the primary factor.
      -Humidity effect is suppressed by temperature as it account for minimal 
       variance on hatchability rates.
       

### Step 3: Post-hoc analysis for individual incubator models(Tukey HSD).


```{r}
tukey_res= TukeyHSD(anova_model,"incubator_model", conf.level= 0.95)
cat("\n=== POST-HOC TUKEY HSD (Incubator model differences)===\n" )
print(tukey_res)
```


NOTE: Incubator model 1 and 4 show a significant difference in hatchability rates from        the rest of the models(2&3).
      
 
      
## Plot Diagnostics.

```{r}
# Anova Residual plots
par(mfrow=c(2,2))
plot(anova_model)
par(mfrow=c(1,1))

```
```{r}
# Box-plot of hatchability by incubator model
boxplot(hatchability_rate~ incubator_model, data= incubator_data,main= "Hatchability Distribution across Incubator Models",xlab="Incubator Model",ylab="Harchability Rate(%)")
```



